---
title: "Baseline Sentiment Survey — Policy Analysis"
author: "Nandni Talreja"
date: "`r Sys.Date()`"
output: html_document
---

## Overview

This script processes baseline survey data collected across 14 battleground states.
For each policy question it produces:

1. **Weighted average allocations** broken down by gender, party, race, age group,
   income group, and education group — both overall and by state
2. **Distribution of opinion** (bucketed into 0–19, 20–39, 40–59, 60–79, 80–100
   support bands) overall and by state
3. **Demographic crosstabs** for income and education subgroups

The analysis uses respondent weights throughout to correct for sample imbalances.

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(readxl)
library(tidyverse)
library(writexl)
```

---

## 1. Load Data

```{r load-data}
# Main survey response file for the focal policy question
# Each row = one respondent's response to one question
df <- read_excel(here::here("data", "survey_raw_policy_a.xlsx"))

# Demographic supplement: income, education, and urban/rural area
# Linked to the main file via User ID
otherdem <- read_excel(here::here("data", "survey_demographics.xlsx"))

# Strip leading/trailing whitespace from all string columns
df <- df %>% mutate(across(where(is.character), str_trim))
otherdem <- otherdem %>% mutate(across(where(is.character), str_trim))

# Target battleground states only
TARGET_STATES <- c("AZ", "CA", "CO", "FL", "ME", "MI", "MO",
                   "MS", "MT", "NE", "NV", "OH", "OR", "WA")

otherdem <- otherdem %>% filter(State %in% TARGET_STATES)

cat("Survey rows:", nrow(df), "\n")
cat("Demo rows:", nrow(otherdem), "\n")
cat("States in demo file:", n_distinct(otherdem$State), "\n")
```

---

## 2. Merge Survey Responses with Demographic Data

```{r merge}
# Merge main responses with demographic supplement by respondent ID
merged <- df %>%
  left_join(otherdem, by = "User ID") %>%
  mutate(user_id_matched = `User ID` %in% otherdem$`User ID`)

# Merge diagnostics — useful for QC when data comes from separate exports
n_total    <- df %>% distinct(`User ID`) %>% nrow()
n_matched  <- df %>% inner_join(otherdem, by = "User ID") %>% distinct(`User ID`) %>% nrow()
n_unmatched <- df %>% anti_join(otherdem, by = "User ID") %>% distinct(`User ID`) %>% nrow()

cat("Total respondents in survey file:", n_total, "\n")
cat("Matched to demographics:", n_matched, "\n")
cat("Unmatched (survey only):", n_unmatched, "\n")
```

---

## 3. Build Demographic Variables

```{r demographics}
# Reshape demographic supplement from long to wide (one row per respondent)
transformeddem <- otherdem %>%
  group_by(`User ID`) %>%
  mutate(row = row_number()) %>%
  pivot_wider(names_from = `Question Name`, values_from = `Option Text`) %>%
  select(-row) %>%
  rename(Area = Area) %>%   # Location urban/suburban/rural already renamed on load
  ungroup()

# Remove respondents who declined to state education (too small to analyze)
transformeddem <- transformeddem %>%
  filter(Education != "Prefer not to say")

# Create binary income group: under vs. over $75k
transformeddem <- transformeddem %>%
  mutate(IncomeGroup = case_when(
    Income %in% c("Under $25,000", "$25,000-$50,000", "$50,000-$75,000") ~ "Under $75,000",
    Income %in% c("$75,000-$100,000", "Over $100,000")                   ~ "Over $75,000"
  ))

# Create binary education group: college vs. non-college
transformeddem <- transformeddem %>%
  mutate(EducationGroup = case_when(
    Education %in% c("College graduate", "Post-graduate school", "Some college") ~ "College",
    Education %in% c("1-11th grade", "High school graduate", "Non-college post HS") ~ "Non-college"
  ))

# Quick summary of demographic distribution
transformeddem %>%
  count(IncomeGroup) %>%
  mutate(Pct = round(100 * n / sum(n), 1))

transformeddem %>%
  count(EducationGroup) %>%
  mutate(Pct = round(100 * n / sum(n), 1))
```

---

## 4. Deduplicate to One Response Per Respondent

```{r deduplicate}
# Some respondents appear multiple times (re-survey or duplicate submissions)
# Keep only the most recent response per respondent
df <- df %>%
  mutate(date = mdy(`Start Date`))

new_df_short <- df %>%
  group_by(`User ID`) %>%
  slice(which.max(as.Date(`Start Date`, "%m/%d/%Y"))) %>%
  ungroup()

# Filter to target states
new_df_short <- new_df_short %>%
  filter(State %in% TARGET_STATES)

# Remove invalid response values (must be 0–100 on the support scale)
new_df_short <- new_df_short %>%
  filter(!is.na(`Response Value`) &
         `Response Value` >= 0 &
         `Response Value` <= 100)

cat("Final deduplicated respondents:", nrow(new_df_short), "\n")
cat("States represented:", n_distinct(new_df_short$State), "\n")
```

---

## 5. Weighted Average Allocations by Demographic Group

```{r weighted-averages}
# Core function: weighted mean of response value within a subgroup
# Uses respondent weight to correct for sample imbalance

# --- Overall + Gender + Party + Race, by state ---
combined <- new_df_short %>%
  group_by(State) %>%
  summarise(
    Overall      = weighted.mean(`Response Value`, `Respondent Weight`),
    Men          = weighted.mean(`Response Value`[Gender == "Male"],
                                 `Respondent Weight`[Gender == "Male"],      na.rm = TRUE),
    Women        = weighted.mean(`Response Value`[Gender == "Female"],
                                 `Respondent Weight`[Gender == "Female"],    na.rm = TRUE),
    Democrats    = weighted.mean(`Response Value`[Party == "Democratic Party"],
                                 `Respondent Weight`[Party == "Democratic Party"], na.rm = TRUE),
    Independents = weighted.mean(`Response Value`[Party == "Independent"],
                                 `Respondent Weight`[Party == "Independent"], na.rm = TRUE),
    Republicans  = weighted.mean(`Response Value`[Party == "Republican Party"],
                                 `Respondent Weight`[Party == "Republican Party"], na.rm = TRUE),
    White        = weighted.mean(`Response Value`[Race == "White"],
                                 `Respondent Weight`[Race == "White"],        na.rm = TRUE),
    `Black/African American` = weighted.mean(
                   `Response Value`[Race == "African American"],
                   `Respondent Weight`[Race == "African American"],           na.rm = TRUE),
    `Native American` = weighted.mean(
                   `Response Value`[Race == "Native American"],
                   `Respondent Weight`[Race == "Native American"],            na.rm = TRUE),
    `Latinx/Hispanic` = weighted.mean(
                   `Response Value`[Race == "Latino / Hispanic"],
                   `Respondent Weight`[Race == "Latino / Hispanic"],          na.rm = TRUE),
    `Asian/AAPI` = weighted.mean(`Response Value`[Race == "Asian"],
                                 `Respondent Weight`[Race == "Asian"],        na.rm = TRUE),
    .groups = "drop"
  )

# --- BIPOC aggregate (all non-White races combined) ---
bipoc_races <- c("African American", "Native American",
                 "Latino / Hispanic", "Asian", "Other")

weighted_mean_bipoc <- new_df_short %>%
  filter(Race %in% bipoc_races) %>%
  group_by(State) %>%
  summarise(BIPOC = weighted.mean(`Response Value`, `Respondent Weight`, na.rm = TRUE),
            .groups = "drop")

# --- Age groups ---
age <- new_df_short %>%
  mutate(AgeGroup = case_when(
    Age >= 18 & Age <= 44 ~ "18-44",
    Age >= 45 & Age <= 64 ~ "45-64",
    Age >= 65              ~ "65+"
  ))

age_groups <- age %>%
  group_by(AgeGroup, State) %>%
  summarise(Avg = weighted.mean(`Response Value`, `Respondent Weight`, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_wider(names_from = AgeGroup, values_from = Avg)

# --- Combine all demographic breakouts into one table ---
alldemos <- combined %>%
  left_join(age_groups,          by = "State") %>%
  left_join(weighted_mean_bipoc, by = "State")

alldemos
```

---

## 6. Distribution of Opinion

```{r dist-of-opinion}
# Bucket responses into 5 equal bands on the 0–100 support scale
# These map to: Strong Oppose / Lean Oppose / Middle / Lean Support / Strong Support

total_weight <- sum(new_df_short$`Respondent Weight`, na.rm = TRUE)

# --- Overall distribution ---
freq_table <- new_df_short %>%
  mutate(Group = case_when(
    `Response Value` >= 0  & `Response Value` <= 19  ~ "0-19",
    `Response Value` >= 20 & `Response Value` <= 39  ~ "20-39",
    `Response Value` >= 40 & `Response Value` <= 59  ~ "40-59",
    `Response Value` >= 60 & `Response Value` <= 79  ~ "60-79",
    `Response Value` >= 80 & `Response Value` <= 100 ~ "80-100"
  )) %>%
  group_by(Group) %>%
  summarise(
    Frequency = (sum(`Respondent Weight`, na.rm = TRUE) / total_weight) * 100,
    .groups = "drop"
  )

print(freq_table)

# --- Distribution by state ---
dist_state <- data.frame(group = c("0-19", "20-39", "40-59", "60-79", "80-100"))

for (st in TARGET_STATES) {
  statewise <- new_df_short %>% filter(State == st)
  total_weight_st <- sum(statewise$`Respondent Weight`, na.rm = TRUE)

  col <- statewise %>%
    mutate(Group = case_when(
      `Response Value` >= 0  & `Response Value` <= 19  ~ "0-19",
      `Response Value` >= 20 & `Response Value` <= 39  ~ "20-39",
      `Response Value` >= 40 & `Response Value` <= 59  ~ "40-59",
      `Response Value` >= 60 & `Response Value` <= 79  ~ "60-79",
      `Response Value` >= 80 & `Response Value` <= 100 ~ "80-100"
    )) %>%
    group_by(Group) %>%
    summarise(
      Frequency = (sum(`Respondent Weight`, na.rm = TRUE) / total_weight_st) * 100,
      .groups = "drop"
    )

  dist_state <- left_join(dist_state, col, by = c("group" = "Group"))
  colnames(dist_state)[ncol(dist_state)] <- st
}

# Reshape to state rows
state_freq_table <- dist_state %>%
  pivot_longer(cols = -group, names_to = "State", values_to = "Frequency") %>%
  pivot_wider(names_from = group, values_from = Frequency)

state_freq_table
```

---

## 7. Demographic Counts (QC)

```{r demo-counts}
# Always verify sample sizes before reporting
# Small subgroups (n < 30) are suppressed in the report and shown as **

age %>%
  group_by(AgeGroup) %>%
  summarise(n = n())

new_df_short %>%
  group_by(Gender) %>%
  summarise(n = n())

new_df_short %>%
  group_by(Party) %>%
  summarise(n = n())

new_df_short %>%
  group_by(State) %>%
  summarise(n = n()) %>%
  arrange(n)

new_df_short %>%
  group_by(State, Race) %>%
  summarise(n = n(), .groups = "drop") %>%
  filter(n < 30) %>%
  arrange(State, Race)
```

---

## 8. Export Results

```{r export}
# Write all output tables to Excel for use in the client report
write_xlsx(
  list(
    "All Demos by State"      = alldemos,
    "Dist of Opinion Overall" = freq_table,
    "Dist of Opinion State"   = state_freq_table
  ),
  here::here("outputs", "policy_a_analysis_results.xlsx")
)

cat("Results exported to outputs/policy_a_analysis_results.xlsx\n")
```
